---
title: "Bayer Leverkusen analysis"
author: "Daryl Dao"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bayer Leverkusen - 2023/24 analysis

## 1. Dataset description

### Source

The dataset used in this report comes from StatsBomb, a well-known advanced sports data provider and analytics. They recently released this dataset to the public after Leverkusen achieved the historic season and the source can be found [here](https://statsbomb.com/articles/soccer/free-statsbomb-data-bayer-leverkusens-invincible-bundesliga-title-win/). The dataset is loaded onto R using StatsBomb's own package called StatsBombR, which is available on [GitHub](https://github.com/statsbomb/StatsBombR) and is installed in the cell below. The matches dataset contains the match information such as teams played, match ID, date that was played, and additional information related to all 34 matches of the season. Using StatsBomb's specific competition and season IDs, all available matches and its accompanied information were first loaded and they were used to load events data from all matches into a single data frame.

### Description

Within the events dataset, it comprises comprehensive event data from all 34 matches that Leverkusen played in this Bundesliga season. Each event is considered as an action that happened when a player interacted with the ball, which was then tagged and categorised into a specific type such as pass, shot, or dribble. This dataset contains key features such as:

-   The information of the match where the event was tagged, including date, minute and second.

-   The information related to the event, including specific ID, event type, player and team who conducted the action, location in x and y coordinates, and duration of the event.

-   Additional information related and specific to each event, including outcome, end location in x and y coordinates, and boolean-type values to indicate whether an event is of a specific type.

These event-related information were used to filter the returned results to only contain relevant events that will be used to plot the visualisations below. This is done to avoid loading irrelevant events as it can be computational demanding to load and store all events at all time.

### **Setup**

```{r load-data, warning=FALSE, message=FALSE, results="hide"}
# Load packages
library(tidyverse) # For loading packages
library(plotly) # For interactive plots
library(dplyr) # For processing data
library(ggplot2) # For visualisation
library(ggsoccer) # For plotting event data

## Install StatsBomb's R package
# devtools::install_github("statsbomb/StatsBombR")
library(StatsBombR)

## Install package to plot football data
# devtools::install_github("jogall/soccermatics")
library(soccermatics)
```

```{r load-statsbomb, warning=FALSE, message=FALSE, results="hide"}
# Load competition information from StatsBomb
BuLi_2324 <- FreeCompetitions() %>% 
  filter(
    competition_id == 9,
    season_id == "281"
  )

# Load all matches from Leverkusen's 23/24 Bundesliga season
BuLi_matches <- FreeMatches(BuLi_2324)

# Load all data
# LeverkusenEvents <- free_allevents(MatchesDF = BuLi_matches, Parallel = T)
# LeverkusenEvents = allclean(LeverkusenEvents)
```

## 2. Visualisations

### Visualisation 1

```{r shots-data, warning=FALSE, message=FALSE, results='hide'}
# Load data
shots_for <- free_allevents(MatchesDF = BuLi_matches, Parallel = T) %>%
  # Use StatsBomb's function to clean the dataframe
  allclean() %>% 
  filter(
    # Select events with type ID of 16 (shot events)
    type.id == 16 &
      # Select events that were made by Leverkusen players
      team.id == 904 &
      # Select relevant shot events and ignore others
      shot.outcome.name %in% c(
        "Off T", "Saved Off Target", "Wayward", "Saved", "Goal"
      )
  ) %>% 
  # Select relevant columns
  select(location.x, location.y, shot.outcome.name, shot.statsbomb_xg)

# Replace specific values
shots_for$shot.outcome.name <- replace(
  shots_for$shot.outcome.name,
  shots_for$shot.outcome.name %in% c("Off T", "Saved Off Target", "Wayward"),
  "Off Target"
)
shots_for$shot.outcome.name <- replace(
  shots_for$shot.outcome.name,
  shots_for$shot.outcome.name == "Saved",
  "On Target"
)
```

```{r shots-for}
ggplot(
  data = shots_for
) +
  # Plot the pitch
  annotate_pitch(
    colour = "#797876",
    fill = "#202020",
    limits = FALSE,
    dimensions = pitch_statsbomb
  ) +
  # Plot the shot points
  geom_point(
    mapping = aes(
      x = location.x,
      y = location.y,
      colour = shot.statsbomb_xg
    )
  ) +
  # Add viridis colourmap
  scale_colour_viridis_c() +
  # Facet by shots type
  facet_grid(~shot.outcome.name) +
  # Use the pitch theme
  theme_pitch() +
  # Change the panel background colour
  theme(
    panel.background = element_rect(fill = "#202020"),
    legend.position = "bottom"
  ) +
  # Set the limits of the x-axis
  coord_flip(
    xlim = c(49, 120)
  ) +
  # Reverse the scale of the y-axis
  scale_y_reverse() +
  # Add labels
  labs(
    colour = "Expected Goals Value",
    title = "Bayer Leverkusen's shots by type and xG",
    subtitle = "From the 2023-24 German Bundesliga season"
  )
```

### Visualisation 2

```{r pass-data, warning=FALSE, message=FALSE, results='hide'}
# Load data
## Passes
passes <- free_allevents(MatchesDF = BuLi_matches, Parallel = T) %>% 
  # Use StatsBomb's function to clean the dataframe
  allclean() %>% 
  filter(
    # Select events with type ID 30 (pass events)
    # and were made by Leverkusen players
    type.id == 30 & team.id == 904
  ) %>% 
  # Select relevant columns
  select(position.name, pass.outcome.name, pass.recipient.name)

## Touches
touches <- free_allevents(MatchesDF = BuLi_matches, Parallel = T) %>%
  # Use StatsBomb's function to clean the dataframe
  allclean() %>%
  # Get all touch events
  # Type ID order: Pass, Ball Receipt, Ball Recovery, Dispossessed, Clearance, Interception, Dribble, Shot, Goal Keeper, Error, Miscontrol, Carry
  filter(
    type.id %in% c(30, 42, 2, 3, 9, 10, 14, 16, 23, 37, 38, 43) & 
      # Player executing the action is from Leverkusen
      # and team in possession when action is executed is Leverkusen
      team.id == 904
  ) %>% 
  # Select relevant columns
  select(location.x, location.y, position.id, position.name, team.name)
```

```{r touches-filter}
# Select relevant columns
touches_filtered <- touches %>% select(
  location.x, location.y, position.id, position.name, team.name
)

# Replace values
# Center Midfield -> Defensive Midfield
touches_filtered$position.id <- replace(
  touches_filtered$position.id,
  touches_filtered$position.id %in% c(15, 13),
  c(11, 9)
)
touches_filtered$position.name <- replace(
  touches_filtered$position.name,
  touches_filtered$position.name %in% c('Left Center Midfield', 'Right Center Midfield'),
  c('Left Defensive Midfield', 'Right Defensive Midfield')
)
# Left/Right Wing -> Left/Right Attacking Midfield
touches_filtered$position.id <- replace(
  touches_filtered$position.id,
  touches_filtered$position.id %in% c(21, 17),
  c(20, 18)
)
touches_filtered$position.name <- replace(
  touches_filtered$position.name,
  touches_filtered$position.name %in% c('Left Wing', 'Right Wing'),
  c('Left Attacking Midfield', 'Right Attacking Midfield')
)

# Retain positions that are in a 3-4-3/3-4-2-1 formation
touches_filtered <- touches_filtered[
  (touches_filtered$position.name %in% c(
    "Goalkeeper", "Left Center Back", "Center Back", "Right Center Back",
    "Left Wing Back", "Right Wing Back", "Left Defensive Midfield",
    "Right Defensive Midfield", "Left Attacking Midfield",
    "Right Attacking Midfield", "Center Forward"
    )),
]
```

```{r avg-positions, warning=FALSE, message=FALSE, results='hide'}
# Calculate the average positions
avg_positions <- touches_filtered %>% 
  # Group the data by position
  group_by(position.id, position.name) %>% 
  # Calculate the mean position
  summarise(
    x = mean(location.x, na.rm = T),
    y = mean(location.y, na.rm = T),
    events = n()
  ) %>% 
  # Remove NA values
  na.omit() %>% 
  # Return a dataframe
  as.data.frame()
```

```{r position-replace}
# Map the pass recipients to their respective main position
passes$pass.recipient.name <- replace(
  passes$pass.recipient.name,
  passes$pass.recipient.name %in% c("Lukáš Hrádecký", "Matěj Kovář"),
  "Goalkeeper"
)
passes$pass.recipient.name <- replace(
  passes$pass.recipient.name,
  passes$pass.recipient.name %in% c("Piero Martín Hincapié Reyna", "Edmond Fayçal Tapsoba"),
  "Left Center Back"
)
passes$pass.recipient.name <- replace(
  passes$pass.recipient.name,
  passes$pass.recipient.name %in% c("Jonathan Tah"),
  "Center Back"
)
passes$pass.recipient.name <- replace(
  passes$pass.recipient.name,
  passes$pass.recipient.name %in% c("Odilon Kossonou"),
  "Right Center Back"
)
passes$pass.recipient.name <- replace(
  passes$pass.recipient.name,
  passes$pass.recipient.name %in% c("Alejandro Grimaldo García", "Arthur Augusto de Matos Soares"),
  "Left Wing Back"
)
passes$pass.recipient.name <- replace(
  passes$pass.recipient.name,
  passes$pass.recipient.name %in% c("Jeremie Frimpong", "Nathan Tella", "Josip Stanišić"),
  "Right Wing Back"
)
passes$pass.recipient.name <- replace(
  passes$pass.recipient.name,
  passes$pass.recipient.name %in% c("Granit Xhaka", "Robert Andrich", "Noah Mbamba", "Nadiem Amiri"),
  "Left Defensive Midfield"
)
passes$pass.recipient.name <- replace(
  passes$pass.recipient.name,
  passes$pass.recipient.name %in% c("Exequiel Alejandro Palacios", "Gustavo Adolfo Puerta Molano"),
  "Right Defensive Midfield"
)
passes$pass.recipient.name <- replace(
  passes$pass.recipient.name,
  passes$pass.recipient.name %in% c("Florian Wirtz", "Adam Hložek"),
  "Left Attacking Midfield"
)
passes$pass.recipient.name <- replace(
  passes$pass.recipient.name,
  passes$pass.recipient.name %in% c("Jonas Hofmann", "Amine Adli"),
  "Right Attacking Midfield"
)
passes$pass.recipient.name <- replace(
  passes$pass.recipient.name,
  passes$pass.recipient.name %in% c("Victor Okoh Boniface", "Patrik Schick", "Borja Iglesias Quintas"),
  "Center Forward"
)
```

```{r pass-filter, warning=FALSE, message=FALSE, results='hide'}
# Filter the passes
passes_filtered <- passes %>% 
  filter(
    # Only retain passes from relevant positions
    position.name %in% avg_positions$position.name
    # Remove passes that are not actually "passes"
    & !(pass.outcome.name %in% c("Injury Clearance", "Pass Offside", "Unknown"))
  ) %>% 
  # Replace outcome NA values with the tag "Successful"
  mutate(
    pass.outcome.name = replace_na(pass.outcome.name, "Successful")
  ) %>% 
  # Group data
  group_by(position.name, pass.recipient.name) %>% 
  # Create a passing matrix using the sum of pass links
  summarise(
    num = n()
  ) %>% 
  # Remove NAs
  na.omit() %>%
  # Only retrieve links that have 100 passes or more
  filter(
    num >= 200
  ) %>% 
  # Return as a data frame
  as.data.frame()
```

```{r edges}
# Join the two dataframes together
edges <- left_join(
  x = passes_filtered, y = avg_positions %>% select(from.x = x, from.y = y, position.name),
  by = "position.name"
)
edges <- left_join(
  x = edges, y = avg_positions %>% select(to.x = x, to.y = y, pass.recipient.name = position.name),
  by = "pass.recipient.name"
)
```

```{r pass-network}
ggplot() +
  # Plot the pitch
  annotate_pitch(
    colour = "#ffffff",
    fill = "#030338",
    limits = TRUE,
    dimensions = pitch_statsbomb,
    linewidth = 0.75
  ) +
  # Plot the pass connections
  geom_segment(
    data = edges,
    mapping = aes(
      x = from.x,
      y = from.y,
      xend = to.x,
      yend = to.y,
      colour = num,
      linewidth = num
    ),
    alpha = 0.7,
    arrow = arrow(length = unit(0.4, "cm")),
    show.legend = c(linewidth = F)
  ) +
  # Add gradient scale for the edges
  scale_colour_gradientn(
    colours = c("#db6576", "#e0df97", "#059f92")
  ) +
  # Plot the average positions
  geom_point(
    data = avg_positions,
    mapping = aes(x = x, y = y),
    fill = "#d7dbda",
    colour = "#e32221",
    size = 5,
    pch = 21
  ) +
  # Add arrow to bottom left to show attacking direction
  annotate(
    geom = "segment",
    x = 0, y = -4,
    xend = 12, yend = -4,
    colour = "white", linewidth = 1,
    arrow = arrow(
      length = unit(0.2, "cm")
    )
  ) +
  # Use theme pitch to remove unnecessary elements
  theme_pitch() +
  # Change the panel background colour
  theme(
    panel.background = element_rect(fill = "#030338"),
    legend.position = "bottom",
    legend.key.width = unit(1.5, "cm")
  ) +
  # Add labels
  labs(
    title = "Bayer Leverkusen's average passing network",
    subtitle = "2023/24 Bundesliga | Min. 200 passes shown",
    colour = "Pass combinations"
  )
```

### Visualisation 3

```{r defActions-data, warning=FALSE, message=FALSE, results='hide'}
# Load data
def_actions <- free_allevents(MatchesDF = BuLi_matches, Parallel = T) %>%
  # Use StatsBomb's function to clean the dataframe
  allclean() %>%
  # Get all defensive action events
  # Type ID order: Ball Recovery, Clearance, Interception, Pressure
  filter(
    type.id %in% c(2, 9, 10, 17) & 
      # Player executing the action is from Leverkusen
      team.id == 904
  ) %>% 
  # Select relevant columns
  select(location.x, location.y, type.id, type.name, player.id, player.name, team.name)
```

```{r defActions-plot}
ggplot() +
  # Plot the pitch
  annotate_pitch(
    limits = TRUE,
    dimensions = pitch_statsbomb,
    linewidth = 0.75
  ) +
  # Plot the heatmap in bins
  stat_bin_2d(
    data = def_actions,
    mapping = aes(x = location.x, y = location.y),
    binwidth = c(10, 10),
    alpha = 0.5,
    na.rm = T
  ) +
  # Add gradient colours
  scale_fill_gradient(low = "white", high = "blue") +
  # Use theme pitch to remove unnecessary elements
  theme_pitch() +
  # Remove legend
  theme(
    panel.background = element_blank(),
    legend.position = "bottom",
    legend.key.width = unit(1.5, "cm")
  ) +
  # Add labels
  labs(
    title = "Bayer Leverkusen | 2023/24 German Bundesliga",
    subtitle = "Defensive actions heatmap",
    fill = "Defensive actions"
  )
```
